---
title: "Metapopulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metapopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

```{r setup}
# Load the requisite packages:
library(malariasimulationGD)
library(malariaEquilibrium)
# Set colour palette:
cols  <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

The metapopulation model runs the individual-based model for multiple parameterized units (countries, regions, admins, etc.) simultaneously. The inputted mixing matrix allows for the transmission of one unit to affect the transmission of another unit. 'Mixing' in the model occurs through the variables `foim_m` and `EIR`. 

Here we will set up a case study of three distinct units and compare output with various transmission mixing patterns. 

## Run metapopulation simulation

### Parameterisation

```{r}
# set variables
year <- 365
human_population <- 1000
sim_length <- 5 * year
EIR_vector <- c(1, 2, 5)

# get parameters
ms_parameterize <- function(x){ # index of EIR
  
  params <- get_parameters(list(human_population = human_population,
                                model_seasonality = FALSE,
                                individual_mosquitoes = FALSE))
  
  # setting treatment
  params <- set_drugs(params, list(AL_params))
  params <- set_clinical_treatment(params, drug = 1, timesteps = 1, coverages = 0.40)

  params <- set_equilibrium(params, init_EIR = EIR_vector[x])
  
  return(params)
  
}

# creating a list of three parameter lists
paramslist <- lapply(seq(1, length(EIR_vector), 1), ms_parameterize)
```

### Simulation 

Our parameters for the three distinct units are stored in the object `paramslist`. Next we will run the metapopulation model with these parameters. We will plug in three mixing matrices - A) isolated, B) semi-mixed, C) perfectly mixed. 

```{r, warning = F}
# isolated
mix_1 <- diag(length(EIR_vector))

# semi-mixed
mix_2 <- matrix(c(0.8, 0.1, 0.1, 
                  0.1, 0.8, 0.1,
                  0.1, 0.1, 0.8),
               nrow = 3, ncol = 3)

# perfectly-mixed
mix_3 <- matrix(rep(1/3, 9), nrow = 3, ncol = 3)

# run model
set.seed(123)

metapop_loop <- function(mixing, mixnam){ # mixing matrix 

  output <- run_metapop_simulation(timesteps = sim_length,
                                   parameters = paramslist,
                                   correlations = NULL,
                                   mixing_tt = 1,
                                   export_mixing = list(mixing), 
                                   import_mixing = list(mixing), 
                                   p_captured = list(matrix(rep(0, 9), nrow = 3, ncol = 3)),
                                   p_captured_tt = 1,
                                   p_success = 0.95)
  
  # convert to dataframe and label EIR and mixing matrix type
  output <-  do.call('rbind', output)
  output$EIR <- c(sort(rep(EIR_vector, sim_length)))
  
  return(output)
  
}

output1 <- metapop_loop(mix_1)
output1$mix <- 'isolated'
  
output2 <- metapop_loop(mix_2)
output2$mix <- 'semi-mixed'
  
output3 <- metapop_loop(mix_3)
output3$mix <- 'perfectly-mixed'
  
output <- rbind(output1, output2, output3)

# get mean PfPR 2-10 by year
output$prev2to10 = output$p_detect_lm_730_3650 / output$n_age_730_3650
output$year = ceiling(output$timestep / 365)
output$mix = factor(output$mix, levels = c('isolated', 'semi-mixed', 'perfectly-mixed'))
output <- aggregate(prev2to10 ~ mix + EIR + year, data = output, FUN = mean)
```

### Explicit overnight human mobility

Explicit human mobility can be used in native metapopulation simulations to
model temporary overnight travel between nodes. This is not permanent migration:
humans remain stored in their home-node records, and `current_node` records
where each human sleeps for exposure and mosquito infection calculations.

The mobility matrix is row-stochastic. Element `human_move_probs[i, j]` is the
probability that a resident of node `i` sleeps in node `j` tonight, conditional
on that resident being at home and available for trip sampling. Diagonal entries
are stay-home probabilities, and off-diagonal entries are travel probabilities.
Rows must sum to 1.

```{r, eval = FALSE}
human_move_probs <- matrix(
  c(
    0.80, 0.20,
    0.10, 0.90
  ),
  nrow = 2,
  byrow = TRUE
)

paramslist <- lapply(seq_len(2), function(node) {
  params <- get_parameters(list(
    human_population = 1000,
    native_mosquito_backend = TRUE,
    individual_mosquitoes = FALSE,
    human_mobility_enabled = TRUE,
    human_trip_duration_type = "fixed",
    human_trip_duration_mean = 1,
    human_mobility_store_diagnostics = TRUE
  ))
  params <- set_equilibrium(params, init_EIR = EIR_vector[node])
  params
})

# The shared matrix can be supplied once. If supplied on multiple nodes,
# all non-NULL matrices must be identical.
paramslist[[1]]$human_move_probs <- human_move_probs

output <- run_metapop_simulation(
  timesteps = sim_length,
  parameters = paramslist,
  correlations = NULL,
  mixing_tt = 1,
  export_mixing = list(diag(2)),
  import_mixing = list(diag(2)),
  p_captured = list(matrix(0, nrow = 2, ncol = 2)),
  p_captured_tt = 1,
  p_success = 0
)

attr(output, "OD_started_trips")
attr(output, "OD_active_overnight_stays")
attr(output, "mean_remaining_trip_duration")
```

With `human_trip_duration_type = "fixed"` and
`human_trip_duration_mean = 1`, off-home trips last one overnight exposure.
Returning travellers spend one timestep at home before possible resampling.

By default, the mobility outputs are `humans_present`, `visitors_present`,
`residents_away`, and `trips_started`. Optional origin-destination diagnostics
are stored only as output-list attributes when
`human_mobility_store_diagnostics = TRUE`.

Explicit mobility currently requires the native metapopulation mosquito backend,
identity import/export mixing, zero border capture, and positive `de`.
`human_move_rates` and non-`"explicit"` mobility modes are unsupported. Native
C++ receives scalar node-level FOIM only; no per-human movement state is passed
to the native backend.

Human-carried state remains attached to the home-node human record, including
bednet state (`net_time`), drugs and prophylaxis (`drug`, `drug_time`), PEV
state, TBV status, immunity, age, `zeta`, `human_slot_contact_multiplier`,
infection state, and infectivity. Bednets are represented through per-human
`net_time`, so they are human-slot attributes. Under explicit mobility v1,
`spray_time` remains attached to the human record. Visitor-specific household
IRS exposure is not explicitly represented. Falciparum blood-immunity boosting
remains tied to the sampled `bitten_humans` pathway.

### Visualisation

Now let's visualize the results of mixing on PfPR2-10:

```{r, fig.align = 'center', out.width='100%', fig.asp = 0.4}
Panel_spec <- rbind(c(0,0.37,0,1),
                    c(0.37,0.635,0,1),
                    c(0.635,0.9,0,1),
                    c(0.9,1,0,1))

Margin_spec <- rbind(c(0.8,0.8,0.3,0.1),
                     c(0.8,0.1,0.3,0.1),
                     c(0.8,0.1,0.3,0.1))


Mix_vec <- unique(output$mix)
EIR_vec <- unique(output$EIR)

Xlab_vec <- c("","Time (years)","")
Ylab_vec <- c(expression(paste(italic(Pf),"PR"[2-10])),"","")
New_vec <- c(F,T,T)

Plot_Mixing <- function(output){
  for(i in 1:3){
    par(fig=Panel_spec[i,], cex = 0.8, mai = Margin_spec[i,], new = New_vec[i])
    with(subset(output,output$mix == Mix_vec[i] & output$EIR == EIR_vec[1]),{
      plot(x = year, y = prev2to10, type = "l", col = cols[1], lwd = 2,
           ylim = c(0,0.36), yaxt = "n",
           xaxs = "i", yaxs = "i",
           ylab = Ylab_vec[i], xlab = Xlab_vec[i])
      abline(h = seq(0.05,0.25, by = 0.05), lty = 2, col = "grey80", lwd = 0.5)
      sapply(2:4, function(x){segments(x0 = x, x1 = x, y0 = 0, y1 = 0.3,
                                       lty = 2, col = "grey80",lwd = 0.5)})
      # grid(lty = 2, col = "grey80", lwd = 0.5)
      if(i ==1){axis(side = 2, at = seq(0,0.4,by=0.1))}
      title(gsub("(^[[:alpha:]])", "\\U\\1", Mix_vec[i], perl=TRUE), line = -1.4, cex.main = 0.9)
      abline(h = 0.3)
      
    })
    for(j in 2:3){
            with(subset(output,output$mix == Mix_vec[i] & output$EIR == EIR_vec[j]),
                 points(x = year, y = prev2to10, type = "l", col = cols[j], lwd = 2))
    }
  }
  par(fig=c(0.9,1,0,1), cex = 0.8, new = T, xpd = T, mai = c(0,0,0,0))
  legend("left", legend = EIR_vec, col = cols[1:3], lwd = 2,
         bty = "n", lty = 1, title = "EIR", y.intersp = 2)
}

Plot_Mixing(output)
```
